\documentclass{article}

\title{BrainSign}

\begin{document}

\maketitle

Submitted to:
DARPA IPTO

Prepared by:
Georgia Institute of Technology


\textbf{Technical Point of Contact:}



\textbf{Administrative Point of Contact:}

\textbf{Co-Principal Investigators:}
Prof. Thad Starner, Georgia Institute of Technology
Prof. Melody Moore Jackson, Georgia Institute of Technology


\begin{enumerate}
 \item DEMONSTRATION SCENARIOS
   \textbf{Scenario 1: Can we recognize single signs in the brain?}

   A person executes a sign while their brain activity is recorded by fMRI. The sequence of movements used to produce the sign resemble a temporal progression of activation in the motor cortex. Previously, fMRI data for common signs were used to estimate a statistical machine learning classifier for sign prediction. This classifier now estimates the sign to which the temporal progression corresponds.

   \textbf{Scenario 2: Can we recognize sign phrases in the brain?}

   A person executes a sign phrase consisting of 4 individual signs while their brain activity is recorded by fMRI. During the signing, the fMRI activation resembles a fading watercolor image with impulses at each sign. Time series activation models previously estimated from a large set of fMRI for various sign phrases are used to classify this sign phrase.

   \item Statement of Work


     \textbf{Task 1 Sign classification from motor cortex activation}
     \textbf{Task 1.1 Single signs}
     
     \paragraph{\underline{Objectives:}}
       Automatically classify single signs from fMRI data of motor cortex activation over time
     \paragraph{\underline{Approach:}}
     Develop a discriminative spatiotemporal model of motor cortex activity during signing. Train a n-class classifier using this representation of the data.
     \paragraph{Milestones:}
     \begin{enumerate}
       \item Month #1: Set the experimental procedures and record fMRI for one subject
       \item Month #5: Construct statistical machine learning models for the fMRI data and rigorously evaluate models via cross-validation
       \item Month #11: Iterate twice with updated experimental procedures from analysis, and evaulate classification accuracy
       \item Month #12: Formalize results and prepare for full proposal
     \end{enumerate}

     \textbf{Task 1.2 Single Phrases}

     \paragraph{\underline{Objectives:}}
       Automatically classify a single phrases from fMRI data of motor cortex activation over time
     \paragraph{\underline{Approach:}}
     Apply state-switching model of neural activation that reuses the single sign model for state-identification, introduces state transition effects related to sign coarticulation, and incorporates sign grammar for the state network's graphical structure. Design a classifier that outputs the most likely sign phrase, given the data.
     \paragraph{Milestones:}
     \begin{enumerate}
       \item Month #1: Set the experimental procedures and record fMRI for one subject
       \item Month #5: Construct statistical machine learning models for the fMRI data and rigorously evaluate models via cross-validation
       \item Month #11: Iterate twice with updated experimental procedures from analysis, and evaulate classification accuracy
       \item Month #12: Formalize results and prepare for full proposal
     \end{enumerate}

     

\end{document}